---
title: "analysing_RNA_seq_with_fsusieR"
author: William Denault and Peter Carbonetto
date: "`r Sys.Date()`"
output:
  html_document:
    toc: no
    highlight: textmate
    theme: readable
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


In this vignette we illustrate how to use fsusieR to fine map *seq data (ATACsaq, RNAseq, CHIPseq).
In this example we focus on RNAseq data, but the process is similar for other *seq data

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Let's first load the data




```{r}
library(fsusieR)
data("RNA_seq_example")
Y=RNAseqData$Y
X=RNAseqData$X
base_pair=RNAseqData$pos
```

This are some synthetic data generated to mimic the RNAseq read in ENSG00000168710 in cortex and the corresponding SNPs effect ont the reads.
the $Y$ matrix contains the RNAseq information of 96 individuals along that ENSG00000168710 spans over
39077 base pairs. The $X$ matrix contains information about the individual genotype level


```{r}
print(dim(Y))
plot( base_pair, Y[1,], main=" Observed base pair count for individual 1")
```


To limit the computational burden we "bin" the data by the dividing the gene body into 
1024 equally large bins. For each individual we sum up the individual counts  over each the bin using
the bin_Y function. The output of the function returns,
** the binned Y data
** The corresponding bin position
** the bon size



```{r}
Y_binned= bin_Y(Y,base_pair = base_pair)
Y=Y_binned$binned_data
print( dim( Y))
pos = Y_binned$pos
print(pos[1:10])
bin_size=Y_binned$bin_size
print( bin_size)
```

Here the corresponding binned count 
```{r}
plot(pos, Y[1,], main=" Observed binned count for individual 1")
```

### Normalization 
Now we want to assert the effect of the genotype of the RNAsea profile
We scale individual count profile by normalizing in a gene-wise fashion as
follow.  Thus allowing to  assess the change of the individual profile 
We asses the effect of the SNP using the $f(x) =log(1+x)$ transform


```{r}
size_factor=rowSums(Y)/ mean(rowSums(Y))
Y_cor =   (  Y /as.vector(size_factor))

plot(pos, log1p(Y[1,] ))
```



### Univariate regression 

the fsusieR package provide a series of function to gauge the effect of genotype of the observed counts.
The simplest function to start with is the **univariate_functional_regression** that allow you to compute the conditional count distribution given a covariate.

Here we look at the effect of SNP 261 on  the observed  log1p counts.


We fit the data using an HMM model, you can also use wavelet re
```{r}
uni_res= univariate_functional_regression(Y=log1p(Y_cor),
                                          X=as.matrix(X[,261],
                                                      ncol=1),
                                          method="smash",
                                          alpha = 0.1
                                          )
```


```{r}
plot(pos, uni_res$effect_estimate, 
     ylab = "estimated effect",
     xlab="base pair",
     type="l",
     lwd=2,
     col="royalblue",
     ylim=c (min(uni_res$cred_band),max(uni_res$cred_band)))
lines( pos,uni_res$cred_band[1,], 
       lty=2,
       lwd=1.4,
       col="royalblue")
lines(pos, uni_res$cred_band[2,], 
        lty=2,
       lwd=1.4,
       col="royalblue")
abline(h=0)

```


### Visualize effect in context
Great you have estimate the effect of a SNP on some count. Let's improve this visualization a bit so we can get some biological interpretation. Below a quick tutorial to put this results in context  using GViz

```{r message=FALSE, warning=FALSE}
library(AnnotationHub)
library(org.Hs.eg.db)
library(GenomicRanges)
library(Gviz)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)


txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
pos0 = pos[1]
pos1= pos[length(pos)]+bin_size
positions=pos
chr=RNAseqData$chr
```


### Create the effect track
 
```{r}
effect=effect=rbind(uni_res$effect_estimate,
             uni_res$cred_band,
             rep(0,length(uni_res$effect_estimate)))
group_colors <- c("black" ,"royalblue","royalblue","royalblue" )
group_lwd= c(1,2,1,1)
x=X[,261]
read_counts <-  rbind( colMeans(log1p(Y_cor[x == 1,])),
                       colMeans(log1p(Y_cor[x == 0,])) 
                           )
obs_effect=rbind (read_counts[2,]-read_counts[1,],
                      rep(0, length(read_counts[2,] )) ) 

ylim= c( min( c(effect,obs_effect)), max(c(  effect,obs_effect)))

group_cred= c(1:3,0)  
effect_track <-
    DataTrack(range = GRanges(seqnames = chr,
                              ranges = IRanges(start = positions,
                                                 end = positions + 1)),
              ylim=ylim,
              data = effect,
              genome = "hg38",
               groups= group_cred,
              name = "Effect SNP 261",
              type = "l",   
              lwd= group_lwd,  
              col = group_colors,
              track.margin = 0.05,
              cex.title = 0.6,
              cex.axis = 0.6,
              col.axis = "black",
              col.title = "black",
              fontface = "plain",
              background.title = "white",
              fontface.title = 1, 
              legend=FALSE)
```


#### Create the gene track

```{r}

  # Create a "gene region" track.
  gene_track <- GeneRegionTrack(txdb,
                                genome = "hg38",
                                chromosome = chr,
                                pos0 = pos0,
                                pos1 = pos1,
                                name = "",
                                showId = TRUE,
                                geneSymbol = TRUE,
                                col.axis = "black",
                                col.title = "black",
                                transcriptAnnotation = "symbol",
                                rotation.title = 0,
                                cex.title = 0.6,
                                col = "salmon",
                                fill = "salmon",
                                background.title = "white")
```


#### Create the observed data point track

```{r}
x       <- X[,261]

read_counts <-  rbind( colMeans(log1p(Y_cor[x == 1,])),
                       colMeans(log1p(Y_cor[x == 0,])) 
                           )
 
groups <- c("SNP 261 = 0",
                "SNP 261 = 1" )
geno_colors <- c("turquoise", "navyblue" )
lab_y = "avg. log1p count" 
  data_track <- DataTrack(range = GRanges(seqnames = chr,
                                          ranges = IRanges(start = positions,
                                                           end = positions + 1)),
                          data = read_counts,
                          genome = "hg38",
                          groups = groups,
                          name = lab_y  , 
                          type = "p" , #"p",#type = "l",
                          col = geno_colors  ,
                          track.margin = 0.05,
                          cex.title = 0.6,
                          cex.axis = 0.6,
                          col.axis = "black",
                          col.title = "black",
                          fontface = "plain",
                          background.title = "white",
                          fontface.title = 1,
                          cex.legend = 0.6,
                          cex=0.2)
```



```{r}
obs_diff_track <-
    DataTrack(range = GRanges(seqnames = chr,
                                          ranges = IRanges(start = positions,
                                                           end = positions + 1)),
                          data =  read_counts[1,]- read_counts[2,],
              genome = "hg38",
              ylim=ylim,             
              name = "observed difference \n between genotype",
              type = "p", 
              lwd= 2,
              col = "blue",
              track.margin = 0.05,
              cex.title = 0.6,
              cex.axis = 0.6,
              col.axis = "black",
              col.title = "black",
              fontface = "plain",
              background.title = "white",
              fontface.title = 1, 
              legend=FALSE)
obs_diff_track2 <-
    DataTrack(range = GRanges(seqnames = chr,
                              ranges = IRanges(start = positions,
                                               end = positions + 1)),
              data =  rep(0, 1024), 
              genome = "hg38",
              ylim= ylim, 
              name ="Observed difference ",
              type = c("l"  ),
              col = c("black" ),
              track.margin = 0.05,
              cex.title =  0.6,
              cex.axis =  0.6,
              col.axis = "black",
              col.title = "black",
              fontface = "plain",
              background.title = "white",
              fontface.title = 1,
              legend = FALSE )
  obs_diff_track =OverlayTrack(trackList = list(obs_diff_track,
                                                  obs_diff_track2),background.title = "white")
  
  
```


Now you can stack the track to visualize them together

```{r}
 tracks <- c( 
      effect_track,
      obs_diff_track,
      data_track,
      gene_track 
    )

plotTracks(tracks, from = pos0, to = pos1, sizes = c(  3,3, 3, 1 ))
```


Ze might have an effect of SNP 261 but it not a given, now let's fine map this data using fsusie

### Fine mapping SNP effect on RNA seq

The example above is nice, however, this naive approach cannot asses if the effect of SNP 261 4
is real or do to linkage disequilbrium with another SNP.

To solve this problem with developed susiF. It can be run as follow. Note that it 
will take a bit of time so you can load the results instead if you are not patient
```{r running_fsusie}
library(fsusieR)
result= susiF(Y=log1p(Y_cor),X=as.matrix(X), L=5, L_start = 5 , post_processing = "smash"   ,
              nullweight = .1,
              thresh_lowcount = .1,
              filter_cs =FALSE)
#result= RNAseqData$results
```


Notice that our candidate in not part of the SNP being selected by susiF



```{r}
result$cs
```





Congrats you have fine map the SNP on our mock RNAseq data. You can visualize the results 
using our custom functions **plot_susiF** to plot both the PIP and the fitted functio, or **plot_susiF_pip** for th PIPS only and **plot_susiF_effect** for visualizing just the effect


 

```{r}
plot_susiF(result
           )
```





```{r}
res_HMM = change_fit   ( result,
                      log1p(Y_cor),
                      X, 
                      to="HMM"
)
```




You can also acces the results directly as follow

```{r}
cs=2 
pos=  result$outing_grid
plot(x= pos ,
     y= result$fitted_func[[cs
                         ]], 
     ylab = paste("estimated effect for CS  ",cs),
     xlab="base pair",
     type="l",
     lwd=2,
     col="royalblue",
     ylim=c (min(result$cred_band[[cs]]),max(result$cred_band[[cs]])))
lines( x= pos ,
       y= result$cred_band[[cs]][1,], 
       lty=2,
       lwd=1.4,
       col="royalblue")
lines(  x= pos ,
        y= result$cred_band[[cs]][2,], 
        lty=2,
       lwd=1.4,
       col="royalblue")
abline(h=0)
```
  
```{r}
tt= do.call( rbind , result$lBF)

plot(apply(tt,2,max), col =ifelse(result$pip>0.2,2,3), pch=20)

```  
We also developed a custom function to visualize the fSuSiE results using GViZ  

You can use the 

 
