---
title: "comparison fSuSiE vs mvSuSiE"
author: "William Denault"
date: "2025-06-24"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{comparison fSuSiE vs mvSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  One SNP difference


In this simulation we showcase that fSuSiE can sometime be much better at
cornering the causal SNP than mvSuSiE when the phenotype is spatially structured



Here we simulate two SNP that are identical except for one individual.

```{r cars}
library(fsusieR)
library(susieR)
set.seed(1)
lev_res= 7
sd_noise=1
n_diff=1
effect_size=0.9
sd_noise=1
# Set parameters
 n <- nrow(N3finemapping$X)         # number of samples
 # Simulate
 X1 <-  N3finemapping$X[ , sample(size=1, 1:ncol(N3finemapping$X))]
X2=X1
val = unique(X1)
  id= sample( size = n_diff, 1:length(X1))
for ( k in id){

  val_id=  which ( !(val==X2[k] ))
   X2[k]= val[ sample(size=1, val_id)]
  }
 # plot(X1,X2)
  X=cbind(X1,X2)
   lf= rep(0,2^lev_res)
  lf[20:30]=effect_size
plot(lf, main="effect of the causal SNP")
  dl= list()
 for ( i in 1:n){
 dl[[i]] = X[i,1]*lf + rnorm (n=2^lev_res, sd=sd_noise)
  }
    Y =do.call(rbind,dl)
```

 Now we run fSuSiE and mvSuSiE to check if they can "tear apart" the correct SNP

```{r  }
library(fsusieR)
res0= susiF(Y=Y,X=X)

library(mvsusieR)
Y_t=Y
prior <- create_mixture_prior(R = ncol(  Y_t))
res= mvsusie(X=X, Y= Y_t, prior_variance = prior)
```

 
 Turns out that SuSiE can but mvSuSiE cannot
```{r}
res$sets

res0$cs
```
 
 
 
