---
title: "Accounting for covariate in functional fine-mapping"
author: "William R.P. Denault"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adjusting_covariate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Adjustement for functional fine-mapping

In order to reduce the number of false positives in fine-mapping analysis, it is often 
important to account for potential counfounders. As most of the tools in the SuSiE suite
don't account for confounders while performing fine-mapping, it is important to  account for potential counfounders prior to the analysis. Similar problem arise with FSuSiE. While it is relatively straightforward to adjust univariate phenotypes. It is more complicated to adjust curves for confounding. Thus we implemented a user freindly function that performs this preprocessing step.


### Generating the data
```{r}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(susiF.alpha)
library(susieR)
library(wavethresh)
set.seed(1)
data(N3finemapping)
attach(N3finemapping)

rsnr <- 1 #expected root signal noise ratio

pos1 <- 250   #Position of the causal covariate for effect 1
pos2 <- 750   #Position of the causal covariate for effect 2
lev_res <- 7#length of the molecular phenotype (2^lev_res)
f1 <-  simu_IBSS_per_level(lev_res )$sim_func#first effect
f2 <- simu_IBSS_per_level(lev_res )$sim_func #second effect

f1_cov  <- simu_IBSS_per_level(lev_res )$sim_func #effect cov 1
f2_cov  <- simu_IBSS_per_level(lev_res )$sim_func #effect cov 2
f3_cov  <- simu_IBSS_per_level(lev_res )$sim_func #effect cov 3

 
```




```{r}

noisy.data  <- list()
Geno <- N3finemapping$X
tt <- svd(Geno)
PCA <- tt$u%*%diag(tt$d)
Cov <- cbind(PCA [,1], PCA [,2], PCA [,3]) 

for ( i in 1:nrow(X))
{
  f1_obs <- f1
  f2_obs <- f2
  noise <- rnorm(length(f1), sd=  (1/  rsnr ) * var(f1))
  noisy.data [[i]] <-  Geno [i,pos1]*f1_obs +Geno [i,pos2]*f2_obs + Cov [i,1]*f1_cov + Cov  [i,2]*f2_cov + Cov  [i,3]*f3_cov +noise

}
noisy.data <- do.call(rbind, noisy.data)

Y <- noisy.data

```


###  Account for covariates

We use the function \textit{adjust_FM_covariate} to regress out the effect of the covariate (in the matrix Cov). The function output an object that contains 
* the adjusted curves
* the fitted effect (covariate)
* the corresponding position for the mapped adjsuted curves and fitted curves

NB: If you input matrix Y has a number of column that is not a power of 2. Then adjust_FM_covariate will remap the corresponding curves to a grid with $2^K$ points. Note that if the corresponding positions of the columns are not evenly spaced then it is important to provide this position in the pos arugment in the pos function. The output of the  The adjusted curves and the fitted effect matrices are mapped on the grid provde


```{r}

out <- adjust_FM_covariate(Y,Cov, pos=1:ncol(Y))
plot( out$Y_adjusted, Y)
plot( out$Y_adjusted, out$Y_fitted)

Y_corrected  <- out$Y_adjusted
```


### Fine-map
We can now fine-map our curves without to be worried of potential confounding due to population stratification or age

```{r, echo=FALSE}
out <- susiF(Y=Y_corrected,
             X=Geno,
             L=3 )


```

We can check the results

 
```{r}
 
 plot_susiF(out)
 

```

### Comparison with fine-mapping without adjustement

We can compare the performance when not adjusting


```{r, echo=FALSE}
out_unadj <- susiF(Y=Y,
             X=Geno,
             L=3 )


```

Here we can the performance 
 
 
```{r}
 par(mfrow=c(1,2))
 
 plot( f1, type="l", main="Estimated effect 1", xlab="")
 lines(get_fitted_effect(out,l=2),col='blue' )
 lines(get_fitted_effect(out_unadj,l=2),col='blue' , lty=2)
 
 abline(a=0,b=0)
 legend(x= 35,
        y=3,
        lty= rep(1,2),
        legend = c("effect 1"," fSuSiE est "),
        col=c("black","blue" )
 )
 plot( f2, type="l", main="Estimated effect 2", xlab="")
 lines(get_fitted_effect(out,l=1),col='green' )
 lines(get_fitted_effect(out_unadj,l=1),col='green' , lty=2)
 
 abline(a=0,b=0)
 legend(x= 20,
        y=-1.5,
        lty= rep(1,2),
        legend = c("effect 2"," fSuSiE est "),
        col=c("black","green" )
 )
 
 par(mfrow=c(1,1))
```
