---
title: "Accounting for covariate in functional fine-mapping"
author: "William R.P. Denault"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accounting for covariate in functional fine-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Adjustement for functional fine-mapping

In order to reduce the number of false positives in fine-mapping
analysis, it is often important to account for potential
counfounders. As most of the tools in the SuSiE suite don't account
for confounders while performing fine-mapping, it is important to
account for counfounders prior to the analysis. Similar problem arise
with fSuSiE. While it is relatively straightforward to adjust
univariate phenotypes. It is more complicated to adjust curves/profile for
confounding. Thus we implemented a user friendly function that
performs this preprocessing step.

### Generating the data
```{r}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(susiF.alpha)
library(susieR)
library(wavethresh)
set.seed(2)
data(N3finemapping)
attach(N3finemapping)

rsnr <- 1 #expected root signal noise ratio

pos1 <- 25   #Position of the causal covariate for effect 1
pos2 <- 75   #Position of the causal covariate for effect 2
lev_res <- 7#length of the molecular phenotype (2^lev_res)
f1 <-  simu_IBSS_per_level(lev_res )$sim_func#first effect
f2 <- simu_IBSS_per_level(lev_res )$sim_func #second effect

f1_cov  <- simu_IBSS_per_level(lev_res )$sim_func #effect cov 1
f2_cov  <- simu_IBSS_per_level(lev_res )$sim_func #effect cov 2
f3_cov  <- simu_IBSS_per_level(lev_res )$sim_func #effect cov 3

 
```


Here the observed data is a mixture of technical noise and genotype signal (target.data). Our goal is to remove the technical noise.

```{r}

Geno <- N3finemapping$X[,1:100]
tt <- svd(N3finemapping$X[,500:700])

Cov <- matrix(rnorm(3*nrow(Geno ),sd=2), ncol=3)
target.data  <-list()
noisy.data  <-list()
for ( i in 1:nrow(Geno))
{
  f1_obs <- f1
  f2_obs <- f2
  noise <- rnorm(length(f1), sd=  (1/  rsnr ) * var(f1))
  target.data [[i]] <-  Geno [i,pos1]*f1_obs +noise +Geno [i,pos2]*f2_obs 
  noisy.data [[ i]] <-   Cov [i,1]*f1_cov  + Cov  [i,2]*f2_cov + Cov  [i,3]*f3_cov

}
technical.noise <- do.call(rbind, noisy.data)
target.data <- do.call(rbind, target.data)
Y <- technical.noise+target.data

```


###  Account for covariates

We use the function \textit{EBmvFR} to regress out the effect of the covariate (in the matrix Cov). The function output an object that contains 
* the adjusted curves
* the fitted effect (covariate)
* the corresponding position for the mapped adjusted curves and fitted curves

NB: If you input matrix Y has a number of column that is not a power of 2. Then \textit{EBmvFR} will remap the corresponding curves to a grid with $2^K$ points. Note that if the corresponding positions of the columns are not evenly spaced then it is important to provide these positions in the pos arugment. The output of the will provide the   adjusted curves and the fitted effect matrices mapped on the grid provided in the pos entry


```{r}

Est_effect <- EBmvFR(Y,X=Cov,adjust=TRUE )
plot(Est_effect$fitted_func[1,],type = "l",
     col="blue",
     lty=2,
     lwd=2)
lines(f1_cov)
legend(x=60,y=-1,
       lwd=c(1,2),
       lty=c(1,2),
       legend=c('effect','estimated'))



Y_corrected  <-Est_effect$Y_adjusted
```


You can also perform the adjustment yourself by accessing the fitted curves directly (and potentially only removing the effect your are interested in)
```{r}
  plot(  Y_corrected, Y-Cov %*%Est_effect$fitted_func )
```




Here you can see that the corrected data are closely matching the actual underlying signals

```{r}

par(mfrow=c(1,2))
plot( Y , target.data)
plot( Y_corrected , target.data)
par(mfrow=c(1,1))
```

### Fine-map
We can now fine-map our curves without to be worried of potential confounding due to population stratification or age

```{r, echo=FALSE}
out <- susiF(Y=Y_corrected  ,
             X=Geno,
             L=10)
```

We can check the results

 
```{r}
 
 plot_susiF(out)
 

```
 
 
  ### Comparison with fine-mapping without adjustement

We can compare the performance when not adjusting


```{r, echo=FALSE}
out_unadj <- susiF(Y=Y,
             X=Geno,
             L=10 )
```

Here we can the performance 



```{r} 
 
 plot( f1, type="l", main="Estimated effect 1", xlab="")
 lines(get_fitted_effect(out,l=2),col='blue' )
 lines(get_fitted_effect(out_unadj,l=2),col='blue' , lty=2)
 
 abline(a=0,b=0)
 legend(x= 20,
        y=-0.5,
        lty= c(1,1,2),
        legend = c("effect 1","fSuSiE ajd", "fSuSiE unajd " ),
        col=c("black","blue","blue" )
 )
 plot( f2, type="l", main="Estimated effect 2", xlab="")
 lines(get_fitted_effect(out,l=1),col='darkgreen' )
 lines(get_fitted_effect(out_unadj,l=1),col='darkgreen' , lty=2)
 
 abline(a=0,b=0)
 legend(x= 10,
        y=2.5,
        lty= c(1,1,2),
        legend = c("effect 1","fSuSiE ajd", "fSuSiE unajd " ),
        col=c("black","darkgreen","darkgreen" )
 )
  
```
