---
title: "fSuSiE applied to binary data"
author: "William Denault"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: no
    highlight: textmate
    theme: readable
vignette: >
  %\VignetteIndexEntry{Introduction to fSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(susieR)
library(fsusieR)
knitr::opts_chunk$set(echo = TRUE)
```

## Generating Binary data

Here we generate $Y$ from a logistic model. Each position as a non zero 
probability of being a 1. Two SNP affect the count over specific region 
the SNP affect a logistic model, so carrying a variant increase the likelihood
of have a 1 at a given position.

We display the effect of the SNP and the baseline below


```{r cars}

set.seed(1)
sigmoid= function(x){
  1/ (1+exp(-x))
}

baseline= rep(-2, 512)
plot(baseline, type="l", ylim=c(-2.2,1.7))
 
effect1= rep(0,512)
 
effect1[100:200]=1
lines(effect1, col="red")
T_m <- 512L
effect2 <- 1.5 * cos(seq_len(T_m) / T_m * 3 * pi)
effect2[effect2 < 0] <- 0
effect2[seq_len(100)] <- 0
```

## Simulate generative model
 
We load some data and sample $Y$ under the model described above
```{r }


library(susieR)
X= N3finemapping$X[1:300, 1:200]
#so having a variant increase the likleihood of getting a 1
for ( j in 1: ncol(X)){

  X[,j]= X[,j]- min(X[,j])
}
true_pos= sample (size=2, 1:ncol(X))

Y= matrix(0, nrow = nrow(X), ncol=512)
for ( i in 1:nrow(Y)){
  prob=sigmoid (baseline+X[i,true_pos[[1]]]*effect1+X[i,true_pos[[2]]]*effect2)
  Y[i,]= rbinom( size=1,prob=prob, n= length(prob) )
} 
image(t(Y), main="Binary Y, row= individual")
```


```{r}
plot( apply(Y,2,mean),

    main= "Average  number of 1s per position")
```

## Application of fSuSiE
As starting point let's just apply fSuSiE with Quantile transform of the
wavelet coefficients. I cut it a 5 iteration for speed purpose but you can let it 
run you will get similar results

```{r message=FALSE}
res_wav= susiF(Y,X, L=10, 
           quantile_trans = TRUE,
           maxit=5 
            )

res_hmm= susiF(Y,X, L=10,post_processing = "HMM",
           quantile_trans = TRUE,
           maxit=5 
            )
res_hmm$cs
res_wav$cs
true_pos

```


You can visualize the estimated effect using plot.
The wavelet fit give you credible band but they can undercover sometimes


```{r}
plot(res_wav)
```


The HMM fit give you point wise local false discovery rate (in black lines ) they
have pretty good lfsr control
```{r}
plot(res_hmm)
```


