---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## System Requirements

### Software Dependencies
- R (≥ 3.5.0)
- Required R packages: wavethresh (≥ 4.6.8), ashr (≥ 2.2-47), mixsqp (≥ 0.3-43), matrixStats (≥ 0.62.0), Rfast (≥ 2.0.5), smashr
- Optional: susieR for comparative analyses

### Operating Systems
Tested on:
- macOS 12.0+
- Ubuntu 20.04+  
- Windows 10+

### Hardware
No special hardware required. Standard desktop computer with 8GB RAM sufficient for typical analyses.

## Installation

Install the latest version from GitHub (typical install time: < 1 minute):

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("stephenslab/smashr")
remotes::install_github("stephenslab/fsusieR")
```

Load the package:

```{r eval=FALSE}
library(fsusieR)
```

## Verify Installation

To verify that the package is working correctly, run the following:

```{r eval=FALSE}
# Load example data
data(N3finemapping)

# Run a quick test
X <- N3finemapping$X[1:50, 1:100]  
Y <- N3finemapping$Y[1:50, 1:128]  
fit <- susiF(X, Y, L = 2)

# Check results
print(fit$cs)
```

If this runs without errors, your installation is successful!