---
title: "Illustration of the limitations of SuSiE for molecular phenotype data"
author: "William R.P. Denault and Gao Wang"
date: "2024-03-07"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustration of the limitations of SuSiE for molecular phenotype data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  results = "hold",
  fig.align = "center",
  dpi = 120,
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
has_mvsusie <- requireNamespace("mvsusieR", quietly = TRUE)
```

This vignette illustrates some of the limitations of SuSiE when
analyzing molecular traits or traits/responses variables that are
functions. We use a simple example where the effect of a SNP spans
over multiple CpG. We start by fine-mapping the effect of the genotype
on each CpG in a sequential fashion. Then, we show that SuSiE can
output inconsistent results. We also look at mvSuSiE under different
prior and residual-variance initializations. Finally we show that
fSuSiE can handle this problem and output much more interpretable
results.

```{r load-pkgs, warning = FALSE, message = FALSE}
library(fsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
```

## Simulation

Let's simulate a simple effect of a SNP on a set of CpGs. The effect
of the SNP is a cosine function that spans over 40 CpGs.

```{r simulate-data, fig.height = 3.25, fig.width = 5}
set.seed(1)
data(N3finemapping, package = "susieR")

X_full <- N3finemapping$X[seq_len(100), ]
T_m <- 128L

effect <- 1.2 * cos(seq_len(T_m) / T_m * 3 * pi)
effect[effect < 0] <- 0
effect[seq_len(40)] <- 0

true_pos <- 700L
y <- matrix(0, nrow = 100L, ncol = T_m)
for (i in seq_len(100L)) {
  y[i, ] <- X_full[i, true_pos] * effect + rnorm(T_m)
}

col_var <- apply(X_full, 2, var)
keep <- which(col_var > 0)
X <- X_full[, keep]
true_pos_filt <- which(keep == true_pos)
affected_pos <- which(effect > 0)

plot(effect, pch = 19, xlab = "CpG", ylab = "effect")
```

We simulate 100 observations of the effect of the SNP on the CpG. The
original causal SNP index is 700. The genotype matrix has a few
constant columns in these 100 samples, so these columns are removed
before fitting the models. This matters for interpreting SNP indices:
the original causal SNP 700 is index 691 in the filtered design matrix.

```{r data-summary}
data.frame(
  N = nrow(X),
  J = ncol(X),
  T = ncol(y),
  true_snp_original = true_pos,
  true_snp_filtered = true_pos_filt,
  affected_start = min(affected_pos),
  affected_end = max(affected_pos),
  affected_positions = length(affected_pos)
)
```

## Method 1: Per-CpG SuSiE

Once the data is simulated, we can run SuSiE on each CpG. For the sake
of simplicity, we only run SuSiE with L = 1 as we are aware that there
is only one effect in this simulation. Also for simplicity, we assume
that the association analysis has been performed with 100%
accuracy—that is, we know exactly which CpGs are the affected CpGs,
and therefore we can ignore the others.

```{r per-position-susie, message = FALSE, warning = FALSE}
susie_est <- vector("list", length(affected_pos))
for (k in seq_along(affected_pos)) {
  susie_est[[k]] <- susie(X, y[, affected_pos[k]], L = 1L)$sets
}
n_cs_per_pos <- vapply(susie_est, function(s) length(s$cs), integer(1L))

knitr::kable(
  data.frame(
    n_credible_sets = as.integer(names(table(n_cs_per_pos))),
    n_positions = as.integer(table(n_cs_per_pos))
  )
)
```

One issue is that SuSiE fails to detect a causal SNP at most of the
affected CpGs:

```{r per-position-plot, fig.height = 4.8, fig.width = 8, message = FALSE, warning = FALSE}
df_effect <- data.frame(position = seq_along(effect), effect = effect)

plot_effect_position <- function(pos) {
  ggplot(df_effect, aes(x = position, y = effect)) +
    geom_point(color = "grey55", size = 1.2) +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
    geom_point(data = df_effect[pos, , drop = FALSE],
               color = "firebrick", size = 2.4) +
    labs(x = "CpG position", y = "true effect",
         title = paste("Position", pos)) +
    theme_classic()
}

plot_susie_position <- function(pos) {
  fit <- susie(X, y[, pos], L = 1L)
  df_pip <- data.frame(snp = keep, pip = fit$pip)
  p <- ggplot(df_pip, aes(x = snp, y = pip)) +
    geom_point(color = "grey55", size = 0.8) +
    geom_vline(xintercept = true_pos, linetype = "dashed",
               color = "firebrick", alpha = 0.7) +
    geom_point(data = df_pip[df_pip$snp == true_pos, , drop = FALSE],
               color = "firebrick", shape = 21, size = 2.3, stroke = 1.1) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = "SNP original index", y = "PIP",
         title = paste("SuSiE PIP at position", pos)) +
    theme_classic()
  cs <- fit$sets$cs$L1
  if (!is.null(cs)) {
    p <- p + geom_point(data = df_pip[cs, , drop = FALSE],
                        color = "dodgerblue", size = 1.8)
  }
  p
}

plot_grid(
  plot_effect_position(88L), plot_effect_position(91L),
  plot_susie_position(88L), plot_susie_position(91L),
  nrow = 2L, ncol = 2L
)
```

A second issue is that the SuSiE produces multiple fine-mapping
signals (and therefore multiple sets of CSs), so it is unclear how to
reason about uncertainty in the causal SNP(s). For example, compare
the SuSiE results for CpGs 88 and 91. At CpG 88, the credible set
misses the true SNP. At CpG 91, the true SNP is present, but it is not
the lead variant.

## Method 2: mvSuSiE

mvSuSiE treats `Y` as a multivariate response. Its behavior depends
strongly on the prior and residual-variance initialization. The
"oracle" residual variance is the identity matrix used to simulate the
data, and the "oracle" prior is the outer product of the true effect
curve with itself. The PIP-zero failure mode of the default mvSuSiE on
this simulation is the subject of
[mvsusieR issue 61](https://github.com/stephenslab/mvsusieR/issues/61).

```{r mvsusie-fits, eval = has_mvsusie, message = FALSE, warning = FALSE}
library(mvsusieR)

U_true <- outer(effect, effect)
prior_true <- create_mixture_prior(
  mixture_prior = list(matrices = list(U_true), weights = 1)
)
prior_canon <- create_mixture_prior(R = T_m)
V_true <- diag(T_m)

run_mvsusie <- function(...) {
  suppressWarnings(mvsusie(..., verbose = FALSE))
}

get_mvsusie_curve <- function(fit) {
  as.numeric(coef(fit)[true_pos_filt + 1L, ])
}

curve_cor <- function(curve) {
  if (sd(curve) == 0) {
    NA_real_
  } else {
    cor(curve, effect)
  }
}

summarize_mvsusie <- function(fit, label, prior, residual, init) {
  curve <- get_mvsusie_curve(fit)
  data.frame(
    test = label,
    prior = prior,
    residual_variance = residual,
    initialization = init,
    pip_at_true_snp = fit$pip[true_pos_filt],
    max_pip = max(fit$pip),
    lead_snp_original = keep[which.max(fit$pip)],
    n_cs = length(fit$sets$cs),
    effect_cor = curve_cor(curve),
    max_abs_effect = max(abs(curve)),
    converged = isTRUE(fit$converged)
  )
}

fit_mv1 <- run_mvsusie(
  X, y, L = 1L,
  prior_variance = prior_true,
  residual_variance = V_true,
  estimate_residual_variance = FALSE,
  estimate_prior_variance = TRUE,
  max_iter = 100L
)

fit_mv2 <- run_mvsusie(
  X, y, L = 1L,
  prior_variance = prior_true,
  residual_variance = V_true,
  estimate_residual_variance = TRUE,
  estimate_prior_variance = TRUE,
  max_iter = 100L
)

fit_mv3 <- run_mvsusie(
  X, y, L = 1L,
  prior_variance = prior_canon,
  residual_variance = V_true,
  estimate_residual_variance = FALSE,
  estimate_prior_variance = TRUE,
  max_iter = 100L
)

fit_mv4 <- run_mvsusie(
  X, y, L = 1L,
  prior_variance = prior_true,
  estimate_residual_variance = TRUE,
  estimate_prior_variance = TRUE,
  max_iter = 100L
)

fit_mv5 <- run_mvsusie(
  X, y, L = 1L,
  prior_variance = prior_true,
  estimate_residual_variance = FALSE,
  estimate_prior_variance = TRUE,
  max_iter = 100L
)

fit_mv6 <- run_mvsusie(
  X, y, L = 1L,
  prior_variance = prior_canon,
  estimate_residual_variance = TRUE,
  estimate_prior_variance = TRUE,
  max_iter = 100L
)

fit_mv7 <- run_mvsusie(
  X, y, L = 1L,
  prior_variance = prior_canon,
  estimate_residual_variance = FALSE,
  estimate_prior_variance = TRUE,
  max_iter = 100L
)

mv_table <- rbind(
  summarize_mvsusie(fit_mv1, "1", "oracle", "oracle fixed",
                    "oracle residual"),
  summarize_mvsusie(fit_mv2, "2", "oracle", "estimated",
                    "oracle residual"),
  summarize_mvsusie(fit_mv3, "3", "canonical", "oracle fixed",
                    "oracle residual"),
  summarize_mvsusie(fit_mv4, "4", "oracle", "estimated", "default"),
  summarize_mvsusie(fit_mv5, "5", "oracle", "default fixed", "default"),
  summarize_mvsusie(fit_mv6, "6", "canonical", "estimated", "default"),
  summarize_mvsusie(fit_mv7, "7", "canonical", "default fixed", "default")
)
```

The tuned configuration recovers the effect shape closely. The default
configuration's recovered curve is essentially flat at zero: with no
truth-informed residual-variance initialization, the PIPs collapse and
the fitted effect carries no information.

```{r mvsusie-effect-curve, eval = has_mvsusie, fig.height = 4, fig.width = 8}
curve_mv_oracle <- get_mvsusie_curve(fit_mv1)
curve_mv_default <- get_mvsusie_curve(fit_mv6)

yrng <- range(c(effect, curve_mv_oracle, curve_mv_default))
plot_mv_curve <- function(curve, main) {
  plot(effect, pch = 20L, col = "grey30", cex = 0.7, ylim = yrng,
       xlab = "position", ylab = "effect", main = main)
  abline(h = 0, lty = 3, col = "grey50")
  lines(curve, col = "firebrick", lwd = 2)
}

par(mfrow = c(1L, 2L), mar = c(4, 4, 2.5, 1))
plot_mv_curve(curve_mv_oracle,
              "mvSuSiE, oracle initialized")
plot_mv_curve(curve_mv_default,
              "mvSuSiE, default")
par(mfrow = c(1L, 1L))
```

The table repeats the same simulation under several configurations:

```{r mvsusie-table, eval = has_mvsusie}
knitr::kable(mv_table, digits = 4)
```

The first three configurations recover the true SNP. All three use the
correct residual-variance initialization. The default initializations
fail: the PIPs collapse to zero and no credible set is returned.

```{r mvsusie-missing, eval = !has_mvsusie, echo = FALSE}
cat("mvSuSiE is not installed, so this comparison is skipped.")
```

## Method 3: fSuSiE

Now compare to an analysis of these data using fSuSiE, which involves
a single call to the "susiF" function. Here we use trend-filtering
post-processing.

```{r run-fsusie, warning = FALSE, message = FALSE}
fit_f <- susiF(y, X, L = 1L, post_processing = "TI", verbose = FALSE)

fit_f$cs
fit_f$pip[true_pos_filt]
```

Notice that fSuSiE returned a single CS containing 1 SNP, which is the
true causal SNP.

```{r plot-fsusie, fig.height = 3.2, fig.width = 8, warning = FALSE, message = FALSE}
df_f_pip <- data.frame(snp = keep, pip = fit_f$pip)
P_pip <- ggplot(df_f_pip, aes(x = snp, y = pip)) +
  geom_point(color = "grey55", size = 0.8) +
  geom_vline(xintercept = true_pos, linetype = "dashed",
             color = "firebrick", alpha = 0.7) +
  geom_point(data = df_f_pip[df_f_pip$snp == true_pos, , drop = FALSE],
             color = "firebrick", size = 2.3) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "SNP original index", y = "PIP", title = "fSuSiE PIPs") +
  theme_classic()

df_f_curve <- data.frame(
  position = seq_along(fit_f$fitted_func[[1]]),
  estimate = fit_f$fitted_func[[1]],
  upper = fit_f$cred_band[[1]][1, ],
  lower = fit_f$cred_band[[1]][2, ],
  truth = effect
)

P_effect <- ggplot(df_f_curve, aes(x = position)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "skyblue",
              alpha = 0.45) +
  geom_point(aes(y = truth), color = "grey30", size = 1.1) +
  geom_line(aes(y = estimate), color = "firebrick", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  labs(x = "CpG position", y = "effect",
       title = "fSuSiE recovered effect curve") +
  theme_classic()

plot_grid(P_pip, P_effect, nrow = 1L, rel_widths = c(1, 1.25))
```

These plots show two other results: on the left-hand side, the
fine-mapping signal (PIPs); on the right-hand side, the estimated
effect on the methylation levels (red = posterior mean, light blue =
credible bands). Together, these are the outputs needed to reason
about both the causal SNP and the shape of its effect across CpGs.
