---
title: "Illustration of the limitations of SuSiE for molecular phenotype data"
author: "William R.P. Denault"
date: "2024-03-07"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustration of the limitations of SuSiE for molecular phenotype data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 3,
  fig.align = "center",
  fig.cap = "&nbsp;",
  dpi = 175
  )
```

# Problem set up 
This vignette illustrates some of the limitations of SuSiE when analyzing molecular traits or traits/responses variables that are functions. We use a simple example where the effect of a SNP spans over multiple CpG. We start by fine-mapping the effect of the genotype on each CpG in a sequential fashion. Then, we show that SuSiE can output inconsistent results. Finally we show that fSuSiE can handle this problem and output much more interpretable results.


Let's simulate a simple effect of a SNP on a set of CpG. The effect of the SNP is a cosine function that spans over 40 CpG. The effect of the SNP is 0 for the first 40 CpG. We simulate 100 observations of the effect of the SNP on the CpG. We then run SuSiE sequentially on each CpG. 

```{r}
 
effect <-   1.2*cos( (1:128)/128 * 3*pi )
effect[which(effect<0)]<- 0
effect[1:40]<- 0

plot( effect, pch=19)
library(fsusieR)
library(susieR)
library(wavethresh)
library(ggplot2)
set.seed(2)
data(N3finemapping)
X <- N3finemapping$X[1:100,]

set.seed(1)

obs <- list()

true_pos <- 700
for (i in 1:100 ){
  obs[[i]] <- X[i,true_pos]*effect+ rnorm(length(effect))
}

y <- do.call(rbind, obs)

which(effect>0)
```
Once the data is simulated, we can run SuSiE on each CpG. For the sake of simplicity, we only run SuSiE with L=1 as we are aware that there is only one effect in this simulation. In the same line, we also limit our fine mapping to the region where the effect is non-zero (which we don't know in practice). 

```{r}
 
susie_est <- list()

for (i in which( effect>0)){
  
  
  susie_est [[i]]<- susie( X=X,y=y[,i], L=1)$sets
  if(!is.null(susie_est[[i]]$cs$L1)){
    print( i)
    print( susie_est [[i]])
  }
}
 
```
SuSiE only detected effect in at  CpGs number  88 and 91. Whereas the effect of the causak SNP (SNP number  700) spans between CpG 65 to CpG 106.Thus showing the lack of power of SuSiE in this scenario.


### Have a look at SuSiE results



```{r echo=FALSE}
df_effect <-data.frame(x=1:128,
                       y=effect)
P00 <- ggplot( df_effect, aes(x=x, y=y))+
  geom_point()+
 
  ylim(c(-0.1,1.3))+
  theme_classic()+
  theme( panel.border = element_rect(colour = "black", fill=NA, size=1.2),axis.text.y=element_blank(),
         axis.ticks.y=element_blank(),
         axis.text.x=element_blank(),
         axis.ticks.x=element_blank())
P01 <- P00+geom_point(x=88,y=effect[88], col="red", shape=21,size=3)+
  
  geom_hline(yintercept = 0)+
  xlab("CpG")+
  ylab("Effect ") 

tt <- susie( X=X,y=y[,88], L=1)

df_pip1 <-data.frame(x=1:length(tt$pip),
                     y=tt$pip)
df2   <-data.frame(x= tt$sets$cs$L1,
                   y=tt$pip[tt$sets$cs$L1])
df3   <-data.frame(x= 700,y=tt$pip[700])

P11 <- ggplot( )+
  geom_point(df_pip1, mapping=aes(x=x, y=y))+
  xlab("SNP index")+
  ylab("PIP")+
  theme_classic()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1.2),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  geom_point(df2 ,  mapping=aes(x=x,y=y), col="lightblue3",size=3)+
  geom_point(df3 ,  mapping=aes(x=x,y=y), col="red", shape=21,size=3)
 



P02 <- P00+geom_point(x=91,y=effect[91], col="red", shape=21,size=3)+
  ylab("")+
  xlab("CpG")+
  ylab("Effect ")+
  geom_hline(yintercept = 0)+
  theme( axis.text.y=element_blank(),
         axis.ticks.y=element_blank(),
         axis.text.x=element_blank(),
         axis.ticks.x=element_blank())
tt <- susie( X=X,y=y[,91], L=1)

df_pip2 <-data.frame(x=1:length(tt$pip),
                     y=tt$pip)
df2   <-data.frame(x= tt$sets$cs$L1,
                   y=tt$pip[tt$sets$cs$L1])
df3   <-data.frame(x= 700,y=tt$pip[700])
P21 <- ggplot( )+
  geom_point(df_pip2, mapping=aes(x=x, y=y))+
  xlab("SNP index")+
  ylab("PIP")+
  theme_classic()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1.2),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  
  geom_point(df2 ,  mapping=aes(x=x,y=y), col="lightblue3",size=3)+
  geom_point(df3 ,  mapping=aes(x=x,y=y), col="red", shape=21,size=3)


P03 <- P00+geom_point(x=94,y=effect[94], col="red", shape=21,size=3)+
  theme_classic()+
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1.2),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  
  geom_hline(yintercept = 0)+
  xlab("CpG")+
  ylab("Effect ") 

 

library(cowplot)
library(gridExtra)
grid.arrange(P01, P11,
             P02,P21,
               ncol=2)

```
 
 
### Let's run fSusiE

```{r}
tt <- susiF(y, X, L=1, post_processing = "smash")
tt$cs
#ten column where removed so we need to remake the plot for the explanation (other t)
tt$pip <- rep( 0,length(tt$pip))
tt$pip[700] <-1

df_pip5 <-data.frame(x=1:length(tt$pip),
                     y=tt$pip)
df_est_f<- data.frame(y=c( tt$fitted_func[[1]],
                           tt$cred_band[[1]][1,],
                           tt$cred_band[[1]][2,]
),
x= rep( 1:128, 3),
type=factor(rep(1:3, each=128)))



df_est_f<- data.frame(y=c( tt$fitted_func[[1]],
                           tt$cred_band[[1]][1,],
                           tt$cred_band[[1]][2,]
),
x= rep( 1:128, 3),
type=factor(rep(1:3, each=128)))


P05 <-  ggplot( )+
  geom_point(df_effect,  mapping=aes(x=x, y=y))+
  geom_line( df_est_f[which(df_est_f$type==1),],
             mapping=aes(x=x, y=y,linetype="longdash"), 
             col="lightblue3",
             size=1.3)+
  geom_line( df_est_f[which(df_est_f$type==2),],
             mapping=aes(x=x, y=y,linetype="solid"),
             col="lightblue3",
             size=1.3)+
  geom_line( df_est_f[which(df_est_f$type==3),],
             mapping=aes(x=x, y=y,linetype="solid"),
             col="lightblue3",
             size=1.3)+
  
  xlab("CpG")+
  ylab("Effect ")+
  theme_classic()+
  geom_hline(yintercept = 0)+
  theme(legend.position = "none",
        panel.border = element_rect(colour = "black", fill=NA, size=1.2),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
P51 <- ggplot( df_pip5, aes(x=x, y=y))+
  geom_point()+
  xlab("SNP index")+
  ylab("PIP")+
  theme_classic()+
  geom_point(x= 700,y=tt$pip[700], col="lightblue3",size=3)+
  theme( panel.border = element_rect(colour = "black", fill=NA, size=1.2),
         axis.text.y=element_blank(),
         axis.ticks.y=element_blank(),
         axis.text.x=element_blank(),
         axis.ticks.x=element_blank())+
  
  geom_point(x= 700,y=tt$pip[700], col="red", shape=21,size=3)
grid.arrange( 
             P05,P51, ncol=2)
```

 
