---
title: "Illustration of the limitations of SuSiE for molecular phenotype data"
author: "William R.P. Denault"
date: "2024-03-07"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustration of the limitations of SuSiE for molecular phenotype data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

This vignette illustrates some of the limitations of SuSiE when
analyzing molecular traits or traits/responses variables that are
functions. We use a simple example where the effect of a SNP spans
over multiple CpG. We start by fine-mapping the effect of the genotype
on each CpG in a sequential fashion. Then, we show that SuSiE can
output inconsistent results. Finally we show that fSuSiE can handle
this problem and output much more interpretable results.

```{r load-pkgs, warning=FALSE, message=FALSE}
library(fsusieR)
library(susieR)
library(wavethresh)
library(ggplot2)
library(cowplot)
```

Let's simulate a simple effect of a SNP on a set of CpGs. The effect
of the SNP is a cosine function that spans over 40 CpGs.

```{r simulate-effect, fig.height=3.25, fig.width=4}
effect <- 1.2 * cos((1:128)/128 * 3*pi)
effect[which(effect < 0)] <- 0
effect[1:40] <- 0
plot(effect,pch = 19)
```

We simulate 100 observations of the effect of the SNP on the CpG.

```{r simulate-data}
set.seed(1)
data(N3finemapping)
X <- N3finemapping$X[1:100,]
y <- matrix(0,100,128)
true_pos <- 700
for (i in 1:100)
  y[i,] <- X[i,true_pos] * effect + rnorm(length(effect))
which(effect > 0)
```

Once the data is simulated, we can run SuSiE on each CpG. For the sake
of simplicity, we only run SuSiE with L = 1 as we are aware that there
is only one effect in this simulation. Also for simplicity, we assume
that the association analysis has been performed with 100%
accuracy—that is, we know exactly which CpGs are the affected CpGs,
and therefore we can ignore the others.

```{r run-susie, message=FALSE, warning=FALSE}
susie_est <- vector("list",128)
for (i in which(effect > 0)) 
  susie_est[[i]] <- susie(X,y[,i],L = 1)$sets
names(susie_est) <- paste0("cpg",1:128)
susie_est <- susie_est[effect > 0]
```

One issue is that SuSiE fails to detect a causal SNP at most of the
affected CpGs:

```{r inspect-susie-results}
sapply(susie_est,function (x) length(x$cs))
```

A second issue is that the SuSiE produces multiple fine-mapping
signals (and therefore multiple sets of CSs), so it is unclear how to
reason about uncertainty in the causal SNP(s). For example, compare
the SuSiE results for CpGs 88 and 91:

```{r plot-susie-res, fig.height=3.5, fig.width=6, warning=FALSE, message=FALSE}
df_effect <- data.frame(x = 1:128,y = effect)
P00 <- ggplot(df_effect,aes(x = x,y = y)) +
  geom_point() +
  ylim(c(-0.1,1.3)) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black",fill = NA,linewidth = 1.2),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
P01 <- P00 +
  geom_point(x = 88,y = effect[88],color = "red",shape = 21,size = 3) +
  geom_hline(yintercept = 0) +
  labs(x = "CpG",y = "effect",title = "CpG 88") 
tt <- susie(X,y[,88],L = 1)
df_pip1 <- data.frame(x = 1:length(tt$pip),y = tt$pip)
df2 <- data.frame(x = tt$sets$cs$L1,y = tt$pip[tt$sets$cs$L1])
df3 <- data.frame(x = 700,y = tt$pip[700])
P11 <- ggplot() +
  geom_point(df_pip1,mapping = aes(x = x,y = y)) +
  labs(x = "SNP",y = "PIP",title = "CpG 88") +
  theme_classic() +
  theme(panel.border = element_rect(color = "black",fill = NA,linewidth = 1.2),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  geom_point(df2,mapping = aes(x = x,y = y),color = "dodgerblue",size = 3)+
  geom_point(df3,mapping = aes(x = x,y = y),color = "red",shape = 21,size = 3)
P02 <- P00 +
  geom_point(x = 91,y = effect[91],color = "red",shape = 21,size = 3) +
  labs(x = "CpG",y = "effect",title = "CpG 91") +
  geom_hline(yintercept = 0) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
tt <- susie(X,y[,91],L = 1)
df_pip2 <- data.frame(x = 1:length(tt$pip),y = tt$pip)
df2 <- data.frame(x = tt$sets$cs$L1,y = tt$pip[tt$sets$cs$L1])
df3 <- data.frame(x = 700,y = tt$pip[700])
P21 <- ggplot() +
  geom_point(df_pip2,mapping = aes(x = x,y = y)) +
  labs(x = "SNP",y = "PIP",title = "CpG 91") +
  theme_classic() +
  theme(panel.border = element_rect(color = "black",fill = NA,linewidth = 1.2),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  geom_point(df2,mapping = aes(x = x,y = y),color = "dodgerblue",size = 3) +
  geom_point(df3,mapping = aes(x = x,y = y),color = "red",shape = 21,size = 3)
P03 <- P00 +
  geom_point(x = 94,y = effect[94],color = "red",shape = 21,size = 3)+
  theme_classic() +
  theme(panel.border = element_rect(color = "black",fill = NA,linewidth = 1.2),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  geom_hline(yintercept = 0) +
  labs(x = "CpG",y = "effect",title = "CpG 91") 
plot_grid(P01,P11,
          P02,P21,
          nrow = 2,ncol = 2)
```
 
Now compare to an analysis of these data using fSuSiE, which involves
a single call to the "susiF" function:

```{r run-fsusie, warning=FALSE, message=FALSE}
tt <- susiF(y,X,L = 1,post_processing = "smash",verbose = FALSE)
tt$cs
```

Notice that fSuSiE returned a single CS containing 1 SNP, which is the
true causal SNP.

These plots show two other results: on the left-hand side, the
fine-mapping signal (PIPs); on the right-hand side, the estimated
effect on the methylation levels (red = posterior mean, light blue =
credible bands).

```{r plot-fsusie-res, fig.height=2, fig.width=6}
tt$pip <- rep( 0,length(tt$pip))
tt$pip[700] <-1
df_pip5 <-data.frame(x=1:length(tt$pip),
                     y=tt$pip)
df_est_f<- data.frame(y = c(tt$fitted_func[[1]],
                            tt$cred_band[[1]][1,],
                            tt$cred_band[[1]][2,]),
                      x = rep(1:128,3),
                      type = factor(rep(1:3,each = 128)))
df_est_f <- data.frame(y = c(tt$fitted_func[[1]],
                             tt$cred_band[[1]][1,],
                             tt$cred_band[[1]][2,]),
                       x = rep(1:128,3),
                       type = factor(rep(1:3,each = 128)))
P50 <- ggplot() +
  geom_point(df_effect,mapping = aes(x = x,y = y)) +
  geom_line(df_est_f[which(df_est_f$type == 1),],
             mapping = aes(x = x,y = y,linetype = "dashed"), 
             color = "red",linewidth = 1) +
  geom_line(df_est_f[which(df_est_f$type == 2),],
            mapping = aes(x = x,y = y),
            color = "skyblue",linewidth = 1)+
  geom_line(df_est_f[which(df_est_f$type == 3),],
            mapping=aes(x = x,y = y),
            color = "skyblue",linewidth = 1) +
  labs(x = "CpG",y = "effect",title = "fSuSiE PIP plot") +
  theme_classic() + 
  geom_hline(yintercept = 0) +
  theme(legend.position = "none",
        panel.border = element_rect(color = "black",fill = NA,linewidth = 1.2),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
P51 <- ggplot(df_pip5,aes(x = x,y = y))+
  geom_point() +
  labs(x = "SNP",y = "PIP",title = "fSusiE effect plot") +
  theme_classic() +
  geom_point(x = 700,y = tt$pip[700],color = "dodgerblue",size = 3)+
  theme(panel.border = element_rect(color = "black",fill = NA,linewidth = 1.2),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  geom_point(x = 700,y = tt$pip[700],color = "red",shape = 21,size = 3)
plot_grid(P51,P50,nrow = 1,ncol = 2)
```
