---
title: "Key ideas for molecular profile QTL fine-mapping with applications to DNAm QTL"
author: William Denault and Peter Carbonetto
date: "`r Sys.Date()`"
output:
  html_document:
    toc: no
    highlight: textmate
    theme: readable
vignette: >
  %\VignetteIndexEntry{Key ideas for molecular profile QTL fine-mapping with applications to DNAm QTL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette we illustrate some of the key ideas underlying fSuSiE
by applying fSuSiE to analyze a toy methylation data set.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in this vignette.

```{r load-pkgs}
library(susieR)
library(fsusieR)
library(reshape2)
library(ggplot2)
library(cowplot)
```

Set the seed so that the results can be reproduced.

```{r set-seed}
set.seed(1)
```

Simulate data
-------------

We simulate a "toy" methylation data set in which the methylation
levels of 100 individuals are measured at 32 CpGs.

Among the 12 candidate SNPs, 3 SNPs affect the methylation levels: 2
SNPs affect the methylation levels of the same cluster of 8 CpGs, and
the other SNP affects the methylation levels of different cluster of 8
CpGs.

```{r sim-params}
n <- 100
m <- 32
p <- 12
```

Generate the SNP minor allele frequencies (MAFs):

```{r sim-mafs}
maf <- 0.05 + 0.45*runif(p)
```

Generate ids for SNPs and CpGs.

```{r sim-ids} 
snpids <- paste0("SNP-",1:p)
cpgids <- paste0("CpG-",1:m)
```

Simulate the genotypes:

```{r sim-geno}
X <- (runif(n*p) < maf) +
     (runif(n*p) < maf)
X <- matrix(X,n,p,byrow = TRUE)
storage.mode(X) <- "double"
X[,4] <- X[,3] + 0.03*rnorm(n)
colnames(X) <- snpids
```

This is the matrix that determines how the SNP alleles change the
methylation levels. 

```{r sim-effect-matrix}
F <- matrix(0,p,m)
F[1,9:16] <- 2.3
F[9,9:16] <- (-2.3)
F[3,25:32] <- 2
rownames(F) <- snpids
colnames(F) <- cpgids
```

Simulate the methylation levels at the CpGs:

```{r sim-Y}
E <- matrix(3*rnorm(n*m),n,m)
Y <- X %*% F
Y <- Y + E
baseline <- min(Y)
Y <- Y - baseline
```

To make the example more realistic, the methylation levels are all
simulated to be zero or higher.

To illustrate, this plot shows the effect of SNP 3 on the methylation
levels. The points show what the data look like and the lines show the
mean methylation levels for the three SNP genotypes:

```{r plot-data, fig.height=2, fig.width=5}
pdat <- melt(Y)
x    <- X[,3]
pdat <- data.frame(cpg        = rep(1:m,each = n) +
                                runif(m*n,min = -0.2,max = 0.2),
                   meth_level = as.vector(Y),
				   geno       = factor(x))
pdat_lines <- data.frame(cpg = rep(1:m,times = 3),
                         geno = factor(rep(0:2,each = m)),
	 			 		 meth_level = rep(c(rep(0,8),rep(-0.75,8),rep(0,16)),
 						                  times = 3))
pdat_lines$meth_level <- pdat_lines$meth_level - baseline
rows <- which(with(pdat_lines,geno == 1 & cpg > 24))
pdat_lines[rows,"meth_level"] <- pdat_lines[rows,"meth_level"] + 2
rows <- which(with(pdat_lines,geno == 2 & cpg > 24))
pdat_lines[rows,"meth_level"] <- pdat_lines[rows,"meth_level"] + 4
p1 <- ggplot(pdat, aes(x = cpg, y = meth_level, color = geno)) +
  geom_point(shape = 20, size = 1.25) +
  scale_x_continuous(breaks = c(0, seq(4, 32, 4))) +
  scale_color_manual(values = c("darkblue", "limegreen", "darkorange")) +
  geom_line(data = pdat_lines, aes(x = cpg, y = meth_level, group = geno), 
            color = "black", linewidth = 1.9) +  # Black line for contrast
  geom_line(data = pdat_lines, linewidth = 1.25) +  # Original colored lines
  labs(x = "CpG", y = "methylation level", color = "genotype") +
  theme_cowplot(font_size = 11)
print(p1)
```

```{r plot-data-ggsave, echo=FALSE, eval=FALSE, message=FALSE}
ggsave("demo_data.pdf",
       p1 + scale_x_continuous(breaks = 1:m),
	   height = 2,width = 5)
```

QTL mapping
-----------

Let's first consider performing association tests for all the CpG-SNP
pairs, which is one approach to mapping methylation QTLs. Here we will
simply use the standard linear regression function in R, `lm()`, to
perform the QTL mapping.

```{r map-qtls}
assoc <- matrix(0,m,p)
rownames(assoc) <- cpgids
colnames(assoc) <- snpids
for (i in 1:m) {
  for (j in 1:p) {
    dat <- data.frame(x = X[,j],y = Y[,i])
    fit <- lm(y ~ x,dat)
    assoc[i,j] <- summary(fit)$coefficients["x","Pr(>|t|)"]
  }
}
```

Having performed these tests for association, we can examine the
associations in two different ways, by SNP and by CpG site. Let's
start with the SNP-centered view:

```{r gwas-by-snp, fig.height=2.5, fig.width=3}
pdat <- data.frame(cpg    = rep(1:m,times = p),
                   snp    = rep(1:p,each = m),
				   effect = as.vector(t(F != 0)),
                   pval   = as.vector(assoc))
pdat <- transform(pdat,pval = -log10(pval))
threshold <- -log10(0.05/(m*p))
p2 <- ggplot(pdat,aes(x = snp,y = pval,shape = effect,color = effect)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = threshold,color = "black",linetype = "dotted") +
  scale_x_continuous(breaks = 1:p) +
  scale_shape_manual(values = c(1,4)) +
  scale_color_manual(values = c("darkblue","darkorange")) +
  labs(x = "SNP",y = "-log10 p-value") +
  theme_cowplot(font_size = 11)
print(p2)
```

Based on these simple association tests, we would identify two out of
the three causal SNPs. (And perhaps we would identify all three causal
SNPs if we were a bit more careful about multiple testing
correction—here we just used the basic Bonferroni correction, which
tells us that only the CpG-SNPs pairs with p-values less than 0.0001
are significant.)

```{r gwas-by-snp-ggsave, echo=FALSE, eval=FALSE}
ggsave("demo_gwas_by_snp.pdf",p2,height = 3,width = 3)
```

*But more importantly*, this view alone doesn't tell us which CPGs are
affected by the SNPs.

This is the CpG-centered view:

```{r gwas-by-cpg, fig.height=2.5, fig.width=3.5}
p3 <- ggplot(pdat,aes(x = cpg,y = pval,shape = effect,color = effect)) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = threshold,linetype = "dotted") +
  scale_x_continuous(breaks = c(0,seq(4,32,4))) +
  scale_shape_manual(values = c(1,4)) +
  scale_color_manual(values = c("darkblue","darkorange")) +
  labs(x = "CpG",y = "-log10 p-value") +
  theme_cowplot(font_size = 11)
print(p3)
```

From the association test p-values alone it is quite clear that the
affected CpG sites "cluster" in two continuous regions, but it is less
clear exactly which CpG sites in these clusters are affected.

```{r gwas-by-cpg-ggsave, echo=FALSE, eval=FALSE, message=FALSE}
ggsave("demo_gwas_by_cpg.pdf",
       p3 + scale_x_continuous(breaks = seq(1,32)),
	   height = 3,width = 3.75)
```

*But more importantly*, this view alone doesn't tell us which SNPs are
affecting the changes in the methylation levels at these CpGs.

Fine-mapping with SuSiE
-----------------------

Let's now consider how one might analyze these data using a standard
(*univariate phenotype*) fine-mapping method, SuSiE, and then we will
point out its limitations and contrast it with fSuSiE.

One possible approach is a two-stage approach where we: (i) identify
the associated CpGs (using the results of the association analysis
above); then (ii) perform fine-mapping separately for each of the
associated CpGs. This is the approach we will try here.

```{r assoc-cpgs}
assoc_cpgs <- subset(pdat,pval > threshold)$cpg
assoc_cpgs <- sort(unique(assoc_cpgs))
assoc_cpgs <- paste0("CpG-",assoc_cpgs)
assoc_cpgs
```

Having identified the 6 associated CpGs, we now run SuSiE separately
for each of them:

```{r run-susie}
num_cpgs <- length(assoc_cpgs)
susie_fits <- vector("list",num_cpgs)
names(susie_fits) <- assoc_cpgs
for (i in assoc_cpgs)
  susie_fits[[i]] <- susie(X,Y[,i],L = 3)
```

This plot shows the 6 fine-mapping "signals" (the PIPs). For convenience,
the two CpG "clusters" are colored red and blue.

```{r pip-plots-susie, fig.height=1.5, fig.width=3}
pdat <- rbind(data.frame(cpg = 9, snp = 1:p,pip = susie_fits[["CpG-9"]]$pip),
              data.frame(cpg = 10,snp = 1:p,pip = susie_fits[["CpG-10"]]$pip),
			  data.frame(cpg = 14,snp = 1:p,pip = susie_fits[["CpG-14"]]$pip),
			  data.frame(cpg = 25,snp = 1:p,pip = susie_fits[["CpG-25"]]$pip),
			  data.frame(cpg = 28,snp = 1:p,pip = susie_fits[["CpG-28"]]$pip),
			  data.frame(cpg = 32,snp = 1:p,pip = susie_fits[["CpG-32"]]$pip))
p5 <- ggplot(pdat,
       aes(x = snp,y = pip,color = factor(cpg),shape = factor(cpg))) +
  geom_point() +
  scale_color_manual(values = c(rep("red",3),rep("darkblue",3))) +
  scale_shape_manual(values = c(1,20,4,1,20,4)) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(breaks = c(0,0.5,1)) +
  labs(x = "SNP",y = "PIP",color = "CpG",shape = "CpG") +
  theme_cowplot(font_size = 10)
p5
```

One takeaway is that it is difficult to interpret these results
because we have 6 fine-mapping signals. fSuSiE, by contrast, is much
easier to interpret result because it produces just a single
fine-mapping signal.

```{r susie-pip-plot-ggsave, echo=FALSE, eval=FALSE}
ggsave("demo_pip_plot_susie.pdf",p5,height = 1.5,width = 3)
```

Upon closer examination, some of the fine-mapping signals agree very
well; for example, SNP 1 is identified as a causal SNP for CpGs 9, 10
and 14 (the red cluster) with a PIP near 1, strongly suggesting that
SNP 1 has an effect on the methylation at all these CpGs. However, the
fine-mapping signal at SNPs 3 and 4 is a little harder to interpret:
one signal (CpG 32) prefers SNP 3, whereas the others prefer SNP 4.
More generally, it is an open question how to integrate these signals
from the separate SuSiE analyses.

Additionally, and perhaps more critically, SuSiE does not directly
tell us which CpGs are affected by each causal SNP. By contrast,
fSuSiE will be able to answer this question.

Fine-mapping with fSuSiE
------------------------

Let's now contrast the QTL mapping and SuSiE analyses with an analysis
of the methylation data using fSuSiE. First we fit the fSuSiE model to
the data. *Notice that we only a fit a single model to all the data*:

```{r run-fsusie, message=FALSE, results="hide"}
fit <- susiF(Y,X,L = 3,filter_cs = FALSE,prior = "mixture_normal",
             post_processing = "HMM")
```

(In this example it is assumed that we know that there are three
causal SNPs, but more generally this number can be estimated.)

fSuSiE correctly produced 3 credible sets (95% CSs) each containing
one of the causal SNPs:

```{r fsusie-cs}
fit$cs
```

This result can be visualized using a "PIP plot" (PIP = posterior
inclusion probability):

```{r fsusie-pip-plot, fig.height=1.5, fig.width=3}
cs_colors <- c("dodgerblue","darkorange","red")
pdat <- data.frame(SNP = 1:p,
                   PIP = fit$pip,
				   CS  = rep("none",p))
pdat$CS[fit$cs[[1]]] <- "L1"
pdat$CS[fit$cs[[2]]] <- "L2"
pdat$CS[fit$cs[[3]]] <- "L3"
pdat <- transform(pdat,CS = factor(CS))
p4 <- ggplot(pdat,aes(x = SNP,y = PIP,fill = CS)) +
  geom_point(shape = 21,size = 2,color = "white") +
  scale_x_continuous(breaks = 1:p) +
  scale_fill_manual(values = c(cs_colors,"gray")) +
  theme_cowplot(font_size = 10)
print(p4)
```

Again, fSuSiE very confidently identified the correct three causal
SNPs, whereas this was not the case in the QTL mapping. The key
difference is that fSuSiE fits a single model to all the data.

```{r fsusie-pip-plot-ggsave, echo=FALSE, eval=FALSE}
ggsave("demo_pip_plot.pdf",p4,height = 1.5,width = 3)
```

fSuSiE also gives a more coherent view of changes to the CpG sites and
how they are affected by the individual causal SNPs (or, more
precisely, by the individual CSs). Here we use the *lfsr* (*local
false sign rate*) estimates produced by the HMM (using the
`post_processing = "HMM"` option):

```{r fsusie-affected-cpgs, fig.height=2, fig.width=6}
i <- sapply(fit$cs,function (x) intersect(x,c(1,3,9)))
pdat <- data.frame(CS   = rep(c("L1","L2","L3"),each = m),
                   CpG  = rep(1:m,times = 3),
                   lfsr = unlist(fit$lfsr_func),
				   affected = as.vector(t(F[i,]) != 0),
				   stringsAsFactors = FALSE)
pdat <- transform(pdat,
                  CS   = factor(CS,c("L3","L2","L1")),
				  lfsr = -log10(lfsr))
ggplot(pdat,aes(x = CpG,y = CS,size = lfsr,color = affected)) +
  geom_point(shape = 1) +
  scale_x_continuous(breaks = c(0,seq(4,32,4))) +
  scale_color_manual(values = c("darkblue","darkorange")) +
  scale_size(range = c(0.5,10),breaks = c(1.3,5,10)) +
  labs(size = "-log10(lfsr)") +
  theme_cowplot(font_size = 10)
```

Indeed, the significance tests (local false sign rate) break down the
support for affected CpGs *separately for each CS* (i.e., for each
causal SNP), and therefore it is quite evident from the fSuSiE results
that two of the causal SNPs affect the same CpG cluster, and the other
causal SNP affects a different CpG cluster.

The lfsrs quantify the support the CpG sites being affected, and in
this example the CpG sites with the smallest lfsrs (largest circles in
the plot) are indeed the affected CpGs.

TI wavelet estimates
--------------------

An alternative to the HMM post-processing step is to use the TI
("translation invariant") wavelets to estimate the effects and their
credible bands:

```{r run-fsusie-TI, message=FALSE, results="hide"}
fit_TI <- susiF(Y,X,L = 3,filter_cs = FALSE,prior = "mixture_normal" )
```

Note that the post-processing step doesn't change the CSs (nor the
PIPs), only the estimated effects of the SNPs on the CpGs.

```{r fsusie-cs-TI}
range(fit$pip - fit_TI$pip)
fit_TI$cs
```

These are the TI wavelet estimates of the effects on the CpG sites
(and corresponding 95% credible bands):

```{r fsusie-affected-cpgs-TI, fig.height=4.5, fig.width=3.5}
out <- lapply(1:3,function (i) get_fitted_effect(fit_TI,l = i,cred_band = TRUE,
			                                     alpha = 0.05))
effect_plot <- function (i) {
  pdat <- data.frame(cpg      = 1:m,
                     estimate = out[[i]]$effect,
					 lower    = out[[i]]$cred_band["low",],
					 upper    = out[[i]]$cred_band["up",])
  rows <- with(pdat,which(lower > 0 | upper < 0))
  pdat2 <- data.frame(cpg = rows,estimate = 0)
  return(ggplot(pdat,aes(x = cpg,y = estimate,ymin = lower,ymax = upper)) +
         geom_point(color = cs_colors[i],size = 1) +
         geom_errorbar(color = cs_colors[i],linewidth = 0.4) +
         geom_hline(yintercept = 0,color = "black",linewidth = 0.4) +
		 geom_point(data = pdat2,mapping = aes(x = cpg,y = estimate),
		            shape = 20,color = "black",size = 1.5,
					inherit.aes = FALSE) +
         scale_x_continuous(breaks = c(0,seq(4,32,4))) +
         labs(x = "CpG",y = "change",title = paste0("CS",i)) +
		 theme_cowplot(font_size = 11))
}
plot_grid(effect_plot(1),
          effect_plot(2),
   	      effect_plot(3),
		  nrow = 3,ncol = 1)
```

The effect estimates with 95% credible bands outside zero predict
very accurately the affected CpG sites.

```{r fsusie-affected-cpgs-TI-ggsave, echo=FALSE, message=FALSE, eval=FALSE}
ggsave("demo_affected_cpgs.pdf",
       plot_grid(effect_plot(1) + scale_x_continuous(breaks = 1:m),
                 effect_plot(2) + scale_x_continuous(breaks = 1:m),
   	             effect_plot(3) + scale_x_continuous(breaks = 1:m),
		  nrow = 3,ncol = 1),
	   height = 4.5,width = 4)
```

Session info
------------

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```
