---
title: "Top-PC SuSiE versus fSuSiE"
author: "William Denault, Peter Carbonetto and Gao Wang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Top-PC SuSiE versus fSuSiE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette compares SuSiE on principal components (PCs) with
fSuSiE. Running SuSiE on PCs is a natural way to reduce a functional
response to one or a few scalar traits. The examples below show two
cases. In the first example, PC 1 captures the simulated signal and
identifies the causal SNP, but it does not estimate the region of the
curve where the SNP acts. In the second example, SuSiE applied to the
first principal component misses signals because different PCs capture
different functional effects.

```{r knitr-opts, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  results = "hold",
  fig.align = "center",
  dpi = 120,
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)
```

```{r load-pkgs, warning = FALSE, message = FALSE}
library(fsusieR)
library(susieR)
library(ggplot2)
library(cowplot)
library(reshape2)
```

## Example 1: one functional effect

This is the same simulation used in the SuSiE limitation vignette. The
effect of one SNP is a cosine function that spans multiple CpGs.

```{r one-signal-sim, fig.height = 3.25, fig.width = 5}
set.seed(1)
data(N3finemapping, package = "susieR")

X_full <- N3finemapping$X[seq_len(100), ]
T_m <- 128L

effect <- 1.2 * cos(seq_len(T_m) / T_m * 3 * pi)
effect[effect < 0] <- 0
effect[seq_len(40)] <- 0

true_pos <- 700L
y <- matrix(0, nrow = 100L, ncol = T_m)
for (i in seq_len(100L)) {
  y[i, ] <- X_full[i, true_pos] * effect + rnorm(T_m)
}

keep <- which(apply(X_full, 2, var) > 0)
X <- X_full[, keep]
true_pos_filt <- which(keep == true_pos)

plot(effect, pch = 19, xlab = "CpG", ylab = "effect")
```

Project the response matrix onto the top PCs and run SuSiE on each PC
score.

```{r one-signal-top-pc, message = FALSE, warning = FALSE}
pca_res <- prcomp(y, center = TRUE, scale. = TRUE)
pve <- pca_res$sdev^2 / sum(pca_res$sdev^2)

pc_results <- data.frame(
  PC = seq_len(5L),
  variance_explained = pve[seq_len(5L)],
  pip_at_true_snp = NA_real_,
  n_cs = NA_integer_,
  lead_snp_original = NA_integer_,
  lead_pip = NA_real_
)

for (pc in seq_len(5L)) {
  fit_pc <- susie(X, pca_res$x[, pc], L = 1L)
  pc_results$pip_at_true_snp[pc] <- fit_pc$pip[true_pos_filt]
  pc_results$n_cs[pc] <- length(fit_pc$sets$cs)
  pc_results$lead_snp_original[pc] <- keep[which.max(fit_pc$pip)]
  pc_results$lead_pip[pc] <- max(fit_pc$pip)
}

knitr::kable(pc_results, digits = 4)
```

PC 1 captures the trait-driven direction cleanly and SuSiE on the PC
1 score returns a single, high-PIP credible set on the true SNP. That
fixes the per-position-SuSiE recall problem: collapsing to one scalar
derived from the curve gives a coherent fine-mapping signal. But PCA
discards the per-position effect direction. The fit tells us a signal
exists at one variant; it does not tell us where on the curve that
variant acts. That target information is what fSuSiE preserves and
PCA loses.

```{r one-signal-fsusie, warning = FALSE, message = FALSE}
fit_one <- NULL
invisible(capture.output({
  fit_one <- susiF(y, X, L = 1L, post_processing = "TI", verbose = FALSE)
}))

fit_one$cs
fit_one$pip[true_pos_filt]
```

```{r one-signal-fsusie-plot, fig.height = 3.2, fig.width = 8}
df_one_pip <- data.frame(snp = keep, pip = fit_one$pip)
P_one_pip <- ggplot(df_one_pip, aes(x = snp, y = pip)) +
  geom_point(color = "grey55", size = 0.8) +
  geom_vline(xintercept = true_pos, linetype = "dashed",
             color = "firebrick", alpha = 0.7) +
  geom_point(data = df_one_pip[df_one_pip$snp == true_pos, , drop = FALSE],
             color = "firebrick", size = 2.3) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "SNP original index", y = "PIP", title = "fSuSiE PIPs") +
  theme_classic()

df_one_curve <- data.frame(
  position = seq_along(fit_one$fitted_func[[1]]),
  estimate = fit_one$fitted_func[[1]],
  upper = fit_one$cred_band[[1]][1, ],
  lower = fit_one$cred_band[[1]][2, ],
  truth = effect
)

P_one_effect <- ggplot(df_one_curve, aes(x = position)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "skyblue",
              alpha = 0.45) +
  geom_point(aes(y = truth), color = "grey30", size = 1.1) +
  geom_line(aes(y = estimate), color = "firebrick", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey50") +
  labs(x = "CpG", y = "effect", title = "fSuSiE recovered effect curve") +
  theme_classic()

plot_grid(P_one_pip, P_one_effect, nrow = 1L, rel_widths = c(1, 1.25))
```

These plots show two results: on the left-hand side, the fine-mapping
signal (PIPs); on the right-hand side, the estimated effect on the
methylation levels (red = posterior mean, light blue = credible
bands).

## Example 2: multiple functional effects

We next regenerate the toy methylation example. The methylation levels
of 100 individuals are measured at 32 CpGs. Among the 12 candidate
SNPs, 3 SNPs affect the methylation levels: 2 SNPs affect the same
cluster of 8 CpGs, and the other SNP affects a different cluster of 8
CpGs.

```{r multi-signal-sim, fig.height = 4.5, fig.width = 8}
set.seed(1)
n <- 100
m <- 32
p <- 12

cs_colors <- c("dodgerblue", "darkorange", "red")

maf <- 0.05 + 0.45 * runif(p)
snpids <- paste0("SNP-", seq_len(p))
cpgids <- paste0("CpG-", seq_len(m))

X <- (runif(n * p) < maf) + (runif(n * p) < maf)
X <- matrix(X, n, p, byrow = TRUE)
storage.mode(X) <- "double"
X[, 4] <- X[, 3] + 0.03 * rnorm(n)
colnames(X) <- snpids

F <- matrix(0, p, m)
F[1, 9:16] <- 2.3
F[9, 9:16] <- -2.3
F[3, 25:32] <- 2
rownames(F) <- snpids
colnames(F) <- cpgids

E <- matrix(3 * rnorm(n * m), n, m)
Y <- X %*% F + E
baseline <- min(Y)
Y <- Y - baseline

x <- X[, 3]
pdat <- data.frame(
  cpg = rep(seq_len(m), each = n) + runif(m * n, min = -0.2, max = 0.2),
  meth_level = as.vector(Y),
  geno = factor(rep(x, times = m))
)
pdat_lines <- data.frame(
  cpg = rep(seq_len(m), times = 3),
  geno = factor(rep(0:2, each = m)),
  meth_level = rep(c(rep(0, 8), rep(-0.75, 8), rep(0, 16)), times = 3)
)
pdat_lines$meth_level <- pdat_lines$meth_level - baseline
rows <- with(pdat_lines, geno == 1 & cpg > 24)
pdat_lines[rows, "meth_level"] <- pdat_lines[rows, "meth_level"] + 2
rows <- with(pdat_lines, geno == 2 & cpg > 24)
pdat_lines[rows, "meth_level"] <- pdat_lines[rows, "meth_level"] + 4

ggplot(pdat, aes(x = cpg, y = meth_level, color = geno)) +
  geom_point(shape = 20, size = 1.25) +
  scale_x_continuous(breaks = c(0, seq(4, 32, 4))) +
  scale_color_manual(values = c("darkblue", "limegreen", "darkorange")) +
  geom_line(data = pdat_lines,
            aes(x = cpg, y = meth_level, group = geno),
            color = "black", linewidth = 1.9) +
  geom_line(data = pdat_lines, linewidth = 1.25) +
  labs(x = "CpG", y = "methylation level", color = "genotype") +
  theme_cowplot(font_size = 11)
```

fSuSiE correctly selects the causal SNPs and outputs estimates that
reflect the effect of the SNP.

```{r multi-signal-fsusie, warning = FALSE, message = FALSE}
fit_multi <- NULL
invisible(capture.output({
  fit_multi <- susiF(
    Y, X, L = 3, filter_cs = FALSE, prior = "mixture_normal",
    post_processing = "TI", verbose = FALSE
  )
}))

fit_multi$cs
```

Visualize the effect.

```{r multi-signal-fsusie-plot, fig.height = 5, fig.width = 7}
out <- lapply(seq_len(3L), function(i) {
  get_fitted_effect(fit_multi, l = i, cred_band = TRUE, alpha = 0.05)
})

effect_plot <- function(i) {
  pdat <- data.frame(
    cpg = seq_len(m),
    estimate = out[[i]]$effect,
    lower = out[[i]]$cred_band["low", ],
    upper = out[[i]]$cred_band["up", ]
  )
  rows <- with(pdat, which(lower > 0 | upper < 0))
  pdat2 <- data.frame(cpg = rows, estimate = 0)
  ggplot(pdat, aes(x = cpg, y = estimate, ymin = lower, ymax = upper)) +
    geom_point(color = cs_colors[i], size = 1) +
    geom_errorbar(color = cs_colors[i], linewidth = 0.4) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    geom_point(data = pdat2, mapping = aes(x = cpg, y = estimate),
               shape = 20, color = "black", size = 1.5,
               inherit.aes = FALSE) +
    scale_x_continuous(breaks = c(0, seq(4, 32, 4))) +
    labs(x = "CpG", y = "change", title = paste0("CS", i)) +
    theme_cowplot(font_size = 11)
}

plot_grid(
  effect_plot(1),
  effect_plot(2),
  effect_plot(3),
  nrow = 3L,
  ncol = 1L
)
```

Now run SuSiE on principal components. Here we use `L = 1` to show
what is captured by each individual PC score.

```{r multi-signal-top-pc, message = FALSE, warning = FALSE}
pca_multi <- prcomp(Y, center = TRUE, scale. = TRUE)
pve_multi <- pca_multi$sdev^2 / sum(pca_multi$sdev^2)
causal_snps <- c(1L, 3L, 9L)

pc_multi_results <- data.frame(
  PC = seq_len(5L),
  variance_explained = pve_multi[seq_len(5L)],
  pip_snp1 = NA_real_,
  pip_snp3 = NA_real_,
  pip_snp9 = NA_real_,
  n_cs = NA_integer_,
  lead_snp = NA_integer_,
  lead_pip = NA_real_
)

pc_multi_fits <- vector("list", 5L)
for (pc in seq_len(5L)) {
  fit_pc <- susie(X, pca_multi$x[, pc], L = 1L)
  pc_multi_fits[[pc]] <- fit_pc
  pc_multi_results$pip_snp1[pc] <- fit_pc$pip[1]
  pc_multi_results$pip_snp3[pc] <- fit_pc$pip[3]
  pc_multi_results$pip_snp9[pc] <- fit_pc$pip[9]
  pc_multi_results$n_cs[pc] <- length(fit_pc$sets$cs)
  pc_multi_results$lead_snp[pc] <- which.max(fit_pc$pip)
  pc_multi_results$lead_pip[pc] <- max(fit_pc$pip)
}

knitr::kable(pc_multi_results, digits = 4)
```

Results on the first PC:

```{r multi-signal-pc1-sets}
pc_multi_fits[[1]]$sets
```

When running SuSiE on the second PC:

```{r multi-signal-pc2-sets}
pc_multi_fits[[2]]$sets
```

In this case, most of the variation in the data is due to orthogonal
effects, and principal components are orthogonal by construction. The
first PC captures the effect of one SNP and has no information about
the other SNP, and vice versa. This is why SuSiE applied to only the
top PC can miss signals.
